Coordinated monitoring of the Piney Point wastewater discharge into Tampa Bay: Data synthesis and reporting

### 2022-11-14 revision/review-response.Rmd

---
output: 
  bookdown::word_document2:
    reference_docx: response_styles.docx
    number_sections: false
always_allow_html: true
urlcolor: blue
link-citations: true
csl: ../florida-scientist.csl
bibliography: ../refs.bib
---

```{r setup, echo = F, warning = F, message = F, results = 'hide'}
# figure path, chunk options
knitr::opts_chunk$set(message = F, echo = F, cache = F, warning = F)
```

```{r echo = F, cache = F, eval = F}
spelling::spell_check_files(here::here('revision/review-response.Rmd'))
```

We sincerely thank the reviewers for providing comments on our manuscript.  We provide a point-by-point response to these comments below. Please note that we have not provided a response to reviewer 3 that did not suggest any changes and recommended publishing the paper as is.

@@

5)  Please list below any suggestions for improvement.

    A brief review for grammar/spelling/punctuation may be warranted. For example, “e.g.” statements should never end in “etc” as the e.g. is intended for examples (e.g., page 6). I imagine that a copy-edit review prior to publication will catch most of these. However, the paper overall was well-written and structured in a logical, easy-to-follow manner.

    Table 2 may need to be reformatted to adjust spacing (hanging ‘a’ in Pennsylvania).

__Response__: Thank you for your comments.  We are pleased you found the manuscript a valuable contribution.  We have proofed the revision for any grammar/spelling/punctuation issues, specifically addressing your comment about "e.g." statements.  The issue noted for Table 2 will be addressed during copy-editing as this is the unformatted version that will be changed on publication.

@@

## Reviewer 2

    Not much “new knowledge” is presented but the paper does a very good job summarizing the events of Piney Point and demonstrating how existing open-science tools can be used by agencies in practice to provide tangible benefits.

    No, this information has not been presented elsewhere. The paper does a good job summarizing how the TBEP’s open science tools allowed for providing information in near real time to provide information to the public and managers, and demonstrates its potential in the wake of the Piney Point discharge, for the TBEP going forward, as well as for other agencies.

    Yes

    No

    A few areas where references could be included, and a better closing sentence emphasizing how much of what was done would not be possible without open-science tools.

__Response__: Thank you for your comments.  We are pleased you noted the importance of open science in delivering these research products to the management community.  Our point-by-point response to the annotated comments provided with your review are below.

Page 2, first paragraph: Maybe add reference to a few case studies or a meta-analysis

* __Response__: The following citations were added: @Tayibi09, @Sanders13, and @Tepanosyan18.

Page 4, first paragraph: Add reference to method

* __Response__: The following citation was added: @Westhoff78

Page 4, third paragraph: "tidy"; use quotes around first mention to show that this is a term specifically from the paper

* __Response__: Quotes added.

Page 9, first paragraph of discussion: Using "However" makes it seem as a counter to the previous sentence. Maybe instead use "For example,"

* __Response__: Sentence revised as follows: "For example, subsequent final closure plans for Piney Point were developed in 2022 following the response-monitoring and open data documentation stemming from this most recent event."

Page 10, final sentence: I think concluding with a remark focusing specifically on how open science provides benefits in these situations would be a better way to end the paper. For example stating how the open-science allows for transparency which increases public trust, made analyses more robust, and facilitated inter-agency cooperation.

* __Response__: A final sentence was added: "Such approaches can facilitate broader public trust and engagement in management decisions by producing more robust, transparent analyses and facilitating inter-agency cooperation."

Figure 2: use semi-colons rather than commas here to separate independent clauses.

* __Response__: Replaced commas with semi-colons.

# References

### 2022-11-14 manu-draft.Rmd
revision done

Human health and environment impacts associated with extractive mining activities are well-documented worldwide.  In Florida, phosphate mining has generated a multi-billion dollar fertilizer industry that supports agricultural production on the global market [@Henderson04]. Seventeen phosphate mining facilities occur in the Tampa Bay watershed and wastewater releases into surface waters from these facilities has negatively impacted environmental resources in the past.  The Piney Point facility on the southeast shore of Tampa Bay has been inactive for over twenty years with the State of Florida overseeing the management of the remaining onsite wastewater. Emergency discharges occurred in the early 2000s and in 2011 to Tampa Bay [@Garrett11;@Switzer11].  In March 2021, a leak was identified in the southern holding pond containment wall and an emergency order by the Florida Department of Environmental Protection (FDEP) authorized wastewater releases to prevent a catastrophic failure of the facility.  Approximately 215 million gallons of mixed legacy phosphate mining wastewater and seawater from port dredging operations were released over a ten day period, introducing ~205 tons of nitrogen directly into lower Tampa Bay [@Beck22].

@@

Water quality monitoring focused on parameters indicative of elevated nutrient loads from the Piney Point release.  Because Tampa Bay is a nitrogen-limited estuary [@Greening14;@Greening06;@Wang99], water quality monitoring focused primarily on total nitrogen and inorganic forms (i.e., total ammonia nitrogen, nitrate/nitrite), chlorophyll-a as an indication of phytoplankton production, and additional parameters expected to respond to nutrient loading (e.g., secchi depth and dissolved oxygen).  Phytoplankton monitoring included both qualitative (presence/absence) and quantitative (cells/L) summaries of taxa enumerated by microscopy from water samples taken at similar frequency and distribution as the water quality sites.  Phytoplankton data also included event-based sampling for *Karenia brevis*.  Finally, seagrass and macroalgae data were collected at transect locations covering the shoreline north of Piney Point and extending to the south into northern Sarasota Bay. Transect sampling included identification of seagrass species and macroalgae, typically to genus, and Braun-Blanquet cover-abundance estimates at 10 m distances along each 50 m transect.  Details on all sampling methods are provided in @Beck22.

@@

A second synthesis challenge was unifying the multiple data types hosted on Google Drive into a usable format for the dashboard and additional analyses (Figure \@ref(fig:synthesis), data synthesis).  Several folders on Google Drive were created that included relevant files for each type of monitoring data (water quality, phytoplankton, etc.).  Each folder typically contained dozens of files saved as Google Sheets, often uploaded as a single sampling event by an individual monitoring agency.  For example, the folder for the water quality sampling included 91 separate files.  Extensive data "wrangling" was required to combine the files into a tidy format [@Wickham14c] and a custom processing script was created by TBEP using the open source R statistical programming language [@RCT20]. An RStudio project was created with full version control that included the synthesis script, `dat_proc.R`, that imported and processed all files from Google Drive into individual binary \.RData files (individual file for water quality, phytoplankton, etc.). These discrete files were later used to develop a subsequent results dashboard .  Several R packages were indispensable in the data synthesis effort and these included the 1) googlesheets4 [@Bryan20] and googledrive [@DAgostino20] R packages for connecting and downloading files from Google Drive, and 2) dplyr, tidyr, lubridate, and readxl packages included with the tidyverse suite of packages [@Wickham19] to assist with data processing.  Importantly, all source files were kept on Google Drive and were unaltered from their original format to isolate the original data from any downstream analysis.

@@

The release of wastewater from Piney Point in the spring of 2021 mobilized resource managers and stakeholders across a variety of institutions in an effort to track and document the potential impacts to Tampa Bay. The communication tools and workflows used to synthesize the data collected in this effort were useful for informing the proactive management response and the analytics of our dashboard suggested that the products created by TBEP were valuable resources that were actively consumed and disseminated in 2021.  However, dashboard analytics are not the only metric of success and additional information can help assess the effectiveness of these tools to efficiently and transparently communicate science for decision-making.  For example, area researchers and managers were able to use screenshots from the dashboard to communicate results in real time, although there was no formal metric for identifying and documenting these applications.  More holistically assessing the value of open science tools is needed, such as how they may be used to inform policy or regulatory changes aimed at minimizing environmental impacts. However, it should be noted that subsequent final closure plans for Piney Point were developed in 2022 following the response-monitoring and open data documentation stemming from this most recent event.

From a data science perspective, several lessons were learned from the need to rapidly synthesize multiple sources of information collected during the response-based monitoring.  Broader use of "tidy" data principles would significantly decrease the processing time required to create data in a usable format for analysis.  Data entry sheets or digital platforms for sharing data should be structured to more easily transform results into a tidy format (i.e., one observation per row, one variable in a column, one value per cell).  However, this requires coordination among multiple entities as to which standards are acceptable for use in digital information storage and even additional training activities to bridge communication gaps between field staff and analysts. Data wrangling will always be a component of analysis, but the time spent doing so could be reduced through routine communication between institutions by adopting similar data standards.  Finding uniform, acceptable solutions for sharing data was also problematic.  Although Google Drive was useful, not every partner was able to access the information given IT limitations that were agency-specific.  Moreover, hosting data on Google Drive was contingent on trust and willingness to share data with TBEP.  The TBEP did not have access to all data collected in response to the 2021 Piney Point event, despite the dozens of partners that did share data with our organization.  Building and sustaining partnerships is a fundamental mission of the National Estuary Programs.  Past successes of TBEP have helped build this trust and the continued success of the organization is dependent on sustaining and growing similar partnerships in the future.

Permanent closure of Piney Point in an environmentally responsible manner remains the ultimate goal of regional resource managers and policy-makers.  The open science tools created by the TBEP in 2021 were effective based on the limited means of assessing their impact, yet it is hoped that the tools developed can continue to facilitate conversations towards broader goals of effectively restoring and protecting Tampa Bay.  In addition to Piney Point, ongoing mining and land development activities in Florida and elsewhere continue to impose external costs on human health and the environment. The role of science in documenting these impacts is important, but even more so, research priorities should focus on understanding societal and political mechanisms that enable these preventable impacts to occur.  Combining traditional science with open science technologies is one path to effectively enable positive change through better science communication.

@@

```{r synthesis, fig.cap = 'Data synthesis workflow for Piney Point data.  Raw data were uploaded by partners to Google Drive, data were synthesized using R-based tools, data and dashboard code were uploaded in a unified format on GitHub that included automated testing, and the dashboard was deployed using Shiny Server.'}
knitr::include_graphics(here('figs/synthesis.JPG'))
```

@@

```{r efftab}
cap.val <- 'Sampling effort for response-based monitoring by partners from April 1st 2021 through September.  Values show the total number of unique sites sampled for water quality, algae (phytoplankton), and seagrass/macroalgae in Tampa Bay.  Values in parentheses show the percentage of the total sites for each monitoring type across months.'
knitr::kable(efftab, booktabs = T, caption = cap.val)
```

>>>>

Human health and environment impacts associated with extractive mining activities are well-documented worldwide [@Tayibi09;@Sanders13;@Tepanosyan18].  In Florida, phosphate mining has generated a multi-billion dollar fertilizer industry that supports agricultural production on the global market [@Henderson04]. Seventeen phosphate mining facilities occur in the Tampa Bay watershed and wastewater releases into surface waters from these facilities has negatively impacted environmental resources in the past.  The Piney Point facility on the southeast shore of Tampa Bay has been inactive for over twenty years with the State of Florida overseeing the management of the remaining onsite wastewater. Emergency discharges occurred in the early 2000s and in 2011 to Tampa Bay [@Garrett11;@Switzer11].  In March 2021, a leak was identified in the southern holding pond containment wall and an emergency order by the Florida Department of Environmental Protection (FDEP) authorized wastewater releases to prevent a catastrophic failure of the facility.  Approximately 215 million gallons of mixed legacy phosphate mining wastewater and seawater from port dredging operations were released over a ten day period, introducing ~205 tons of nitrogen directly into lower Tampa Bay [@Beck22].

@@

Water quality monitoring focused on parameters indicative of elevated nutrient loads from the Piney Point release.  Because Tampa Bay is a nitrogen-limited estuary [@Greening14;@Greening06;@Wang99], water quality monitoring focused primarily on total nitrogen and inorganic forms (i.e., total ammonia nitrogen, nitrate/nitrite), chlorophyll-a as an indication of phytoplankton production, and additional parameters expected to respond to nutrient loading (e.g., secchi depth and dissolved oxygen).  Phytoplankton monitoring included both qualitative (presence/absence) and quantitative (cells/L) summaries of taxa enumerated by microscopy from water samples taken at similar frequency and distribution as the water quality sites.  Phytoplankton data also included event-based sampling for *Karenia brevis*.  Finally, seagrass and macroalgae data were collected at transect locations covering the shoreline north of Piney Point and extending to the south into northern Sarasota Bay. Transect sampling included identification of seagrass species and macroalgae, typically to genus, and Braun-Blanquet cover-abundance estimates [@Westhoff78] at 10 m distances along each 50 m transect.  Details on all sampling methods are provided in @Beck22.

@@

A second synthesis challenge was unifying the multiple data types hosted on Google Drive into a usable format for the dashboard and additional analyses (Figure \@ref(fig:synthesis), data synthesis).  Several folders on Google Drive were created that included relevant files for each type of monitoring data (water quality, phytoplankton, etc.).  Each folder typically contained dozens of files saved as Google Sheets, often uploaded as a single sampling event by an individual monitoring agency.  For example, the folder for the water quality sampling included 91 separate files.  Extensive data "wrangling" was required to combine the files into a "tidy" format [@Wickham14c] and a custom processing script was created by TBEP using the open source R statistical programming language [@RCT20]. An RStudio project was created with full version control that included the synthesis script, `dat_proc.R`, that imported and processed all files from Google Drive into individual binary \.RData files (individual file for water quality, phytoplankton, etc.). These discrete files were later used to develop a subsequent results dashboard .  Several R packages were indispensable in the data synthesis effort and these included the 1) googlesheets4 [@Bryan20] and googledrive [@DAgostino20] R packages for connecting and downloading files from Google Drive, and 2) dplyr, tidyr, lubridate, and readxl packages included with the tidyverse suite of packages [@Wickham19] to assist with data processing.  Importantly, all source files were kept on Google Drive and were unaltered from their original format to isolate the original data from any downstream analysis.

@@

The release of wastewater from Piney Point in the spring of 2021 mobilized resource managers and stakeholders across a variety of institutions in an effort to track and document the potential impacts to Tampa Bay. The communication tools and workflows used to synthesize the data collected in this effort were useful for informing the proactive management response and the analytics of our dashboard suggested that the products created by TBEP were valuable resources that were actively consumed and disseminated in 2021.  However, dashboard analytics are not the only metric of success and additional information can help assess the effectiveness of these tools to efficiently and transparently communicate science for decision-making.  For example, area researchers and managers were able to use screenshots from the dashboard to communicate results in real time, although there was no formal metric for identifying and documenting these applications.  More holistically assessing the value of open science tools is needed, such as how they may be used to inform policy or regulatory changes aimed at minimizing environmental impacts. For example, subsequent final closure plans for Piney Point were developed in 2022 following the response-monitoring and open data documentation stemming from this most recent event.

From a data science perspective, several lessons were learned from the need to rapidly synthesize multiple sources of information collected during the response-based monitoring.  Broader use of tidy data principles would significantly decrease the processing time required to create data in a usable format for analysis.  Data entry sheets or digital platforms for sharing data should be structured to more easily transform results into a tidy format (i.e., one observation per row, one variable in a column, one value per cell).  However, this requires coordination among multiple entities as to which standards are acceptable for use in digital information storage and even additional training activities to bridge communication gaps between field staff and analysts. Data wrangling will always be a component of analysis, but the time spent doing so could be reduced through routine communication between institutions by adopting similar data standards.  Finding uniform, acceptable solutions for sharing data was also problematic.  Although Google Drive was useful, not every partner was able to access the information given IT limitations that were agency-specific.  Moreover, hosting data on Google Drive was contingent on trust and willingness to share data with TBEP.  The TBEP did not have access to all data collected in response to the 2021 Piney Point event, despite the dozens of partners that did share data with our organization.  Building and sustaining partnerships is a fundamental mission of the National Estuary Programs.  Past successes of TBEP have helped build this trust and the continued success of the organization is dependent on sustaining and growing similar partnerships in the future.

Permanent closure of Piney Point in an environmentally responsible manner remains the ultimate goal of regional resource managers and policy-makers.  The open science tools created by the TBEP in 2021 were effective based on the limited means of assessing their impact, yet it is hoped that the tools developed can continue to facilitate conversations towards broader goals of effectively restoring and protecting Tampa Bay.  In addition to Piney Point, ongoing mining and land development activities in Florida and elsewhere continue to impose external costs on human health and the environment. The role of science in documenting these impacts is important, but even more so, research priorities should focus on understanding societal and political mechanisms that enable these preventable impacts to occur.  Combining traditional science with open science technologies is one path to effectively enable positive change through better science communication. Such approaches can facilitate broader public trust and engagement in management decisions by producing more robust, transparent analyses and facilitating inter-agency cooperation.

@@

```{r synthesis, fig.cap = 'Data synthesis workflow for Piney Point data.  Raw data were uploaded by partners to Google Drive; data were synthesized using R-based tools; data and dashboard code were uploaded in a unified format on GitHub that included automated testing; and the dashboard was deployed using Shiny Server.'}
knitr::include_graphics(here('figs/synthesis.JPG'))
```

@@

```{r efftab}
cap.val <- 'Sampling effort for response-based monitoring by partners from April 1st, 2021 through September.  Values show the total number of unique sites sampled for water quality, algae (phytoplankton), and seagrass/macroalgae in Tampa Bay.  Values in parentheses show the percentage of the total sites for each monitoring type across months.'
knitr::kable(efftab, booktabs = T, caption = cap.val)
```
